Co-citations in context

### 2019-06-10 cocit_r.tex
rounding off merge

\citep{uzzi_atypical_2013} used a novel approach for co-citation analysis using, as input, 17.9 million articles and their cited references, from the Web of Science (WoS)  to characterize a subset of highly cited articles with respect to both novel and conventional combinations of prior research. In this study, the frequency with which references were calculated and reduced to journal pairs frequencies (observed co-citation frequencies). Expected values were generated from Monte Carlo simulations under a random graph model. Observed frequencies were then normalized (shifted and scaled) to averaged expected values from ten simulations and termed as \emph{z-scores}, which were interpreted as indicating conventional thought if positive and novel thought if negative. Consequently, every article was associated with multiple z-scores corresponding to co-cited journal pairs in its refernces. For each article, positional statistics of z-scores were calculated to set thresholds for a binary (high/low) classification of conventionality (C) using the median and novelty (N) using the tenth percentile of z-scores within an article. Thus, HCLN would denote high conventionality and low novelty with all four combinations being possible. 
\cite{uzzi_atypical_2013} observed that HCHN articles were twice as likely to be highly cited, suggesting that novel combinations of ideas flavoring a body of conventional thought were a 'nearly universal` feature for impact.

@@

A follow-up study by \citep{boyack_vs_uzzi_2014} explored the impact of discipline and journal effects on the definition of conventionality and novelty.  While their study had some methodological differences in the use of Scopus data rather than WoS data, a smaller dataset, and a $\chi^2$ calculation rather than Monte Carlo simulations to generate expected values of journal pairs, Boyack and Klavans also reported the trend that HCHN is more probable in highly cited papers. However, they note that ``only 64.4\%  of  243  WoS  subject  categories'' in the Uzzi {\em et al.} study met the criterion of having the highest probability of hit papers in the HCHN category.  Further, they observed that journals vary widely in terms of size and influence and that 20 journals accounted for nearly 15\% of co-citations in their measurements. Lastly, they noted that three multidisciplinary journals accounted for 9.4\% of all atypical combinations, suggesting strong effects from both disciplines and journals that were not reported by Uzzi et al.

Despite different methods used to generate expected values, both of these key preceding studies (\emph{vide supra}) measured co-citation frequencies across the scientific literature without disciplinary constraints and subsequently analyzed disciplinary subsets. In extending their prior work, we hypothesized that normalization using expected values drawn from datasets focused in disciplinary areas, rather than all of WoS, would reduce model mis-specification.  Consequently, we used keyword searches of the scientific literature to construct citation networks themed around academic disciplines: \emph{applied physics, immunology, and metabolism}. Within these disciplinary frameworks, we calculated observed and expected co-citation frequencies using a refined random graph model and an efficient Monte Carlo simulation algorithm.

@@

\section{Materials and Methods}
\emph{Bibliographic data} We have previously developed ERNIE, an open source knowledge platform into which we parse the Web of Science (WoS) Core Collection~\citep{Keserci371955}. WoS data stored in ERNIE spans the period 1900-2019 and consists of over 72 million publications. For this study, we generated an analytical dataset from years 1985 to 2005 from ERNIE. The total number of publications in this dataset was just over 25 million publications (25,134,073), which were then stratified by year of publication. For each of these years, we further restricted analysis to publications of type Article. Since WoS data also contains incomplete references or references that point at other indexes, we also considered only those references for which there were complete records~(Table \ref{tab:summary_data}). For example, WoS data for year 2005 contained 1,753,174 publications, which after restricting to type Article and considering only those references described above resulted in 916,573 publications, 6,095,594 unique references (set of references), and 17,167,347 total references (multiset of references). Given consistent trends in the data, Table \ref{tab:summary_data}, we analyzed the two boundary years (1985 and 2005) and the mid-point (1995) performing 1,000 simulations for each dataset. We also calculated the number of times each of these articles was cited in the first 8 years since publication.

@@

\subsection{Model Misspecification and the Attributes of Disciplinary Context} We invoke model mis-specification arising from irrelevant references as a concern when disciplinary subsets are analyzed using a random graph model that generates expected co-citation frequencies from the full body of scientific literature. To address this issue, we first assembled three exemplar datasets using keyword searches for the terms `applied physics', `immunology', and `metabolism' (Materials and Methods). Using the WoS extended classificatation of 153 subject areas, we then analyzed the disciplinary composition of the references in these three datasets into the five research areas in the Web of Science (life sciences and biomedicine, physical sciences, social sciences, technology, and arts and humanities) for the years 1985, 1995, and 2005. We observed that 84.2\% of the references in the applied physics dataset were classified as belonging to the WoS physical sciences research area. Similarly, 93.7\% and 92.1\% of the references in the immunology and metabolism datasets belonged to the life sciences and biomedicine research area, indicating strongly that references are more likely belong to the same research area as the publications that cite them.

A symptom of model misspecification is that random sampling of citations associated with an article do not match with the distribution of citations across the disciplines observed in articles from that discipline. \cite{wallace_lariviere_gingras_2012} found that, often, a majority of an article's citations are from the specialty of the article, although that percentage varied among disciplines in the eight specialties they investigated: from approximately 39\% to 89\% for 2006. We also analyzed the consequences of citation shuffling within a disciplinary set or all of the Web of Science.  References in publications belonging to these three datasets were classified using the WoS Extended Subjects classification of 153 subject areas and summarized as a frequency distribution. A single shuffle of the references in these datasets or all the references in the corresponding WoS year slice was performed, using either the rpcs or umsj algorithms, after which subject frequencies were computed again. The fold change in subject frequencies of references before and after shuffling was calculated for these groups using all 153 Subject categories and are summarized as boxplots in Fig \ref{fig:be}. For example, the applied physics dataset contained 1 reference labeled Genetics and Heredity . This proportion was preserved in the intra-network shuffle but increased 1496 times when the WoS background was used as a source of substitution during Monte Carlo simulation with the repcs algorithm. Similarly, with the metabolism dataset.which contained one reference labeled Philosophy that was increased to 661 after a single repcs shuffle. The data show convincingly that the disciplinary composition of references in a network is preserved when citation shuffling is conducted within it but is significantly altered when the full WoS network is used as a source of substitution. A second inference is that the choice of algorithm is not relevant beyond runtime advantage but that the background network used for the shuffle is critical.

>>>>

\citep{uzzi_atypical_2013} used a novel approach for co-citation analysis, using 17.9 million articles and their cited references from the Web of Science (WoS) as input, to characterize a subset of highly cited articles with respect to both novel and conventional combinations of prior research. The frequency with which references were co-cited was calculated and expressed as journal pair frequencies (observed co-citation frequencies). Expected values were generated from Monte Carlo simulations under a random graph model. Observed frequencies were then normalized (shifted and scaled) to averaged expected values from ten simulations and termed as \emph{z-scores}, which were interpreted as indicating conventional thought if positive and novel thought if negative. Consequently, every article was associated with multiple z-scores corresponding to co-cited journal pairs in its refernces. For each article, positional statistics of z-scores were calculated to set thresholds for a binary (high/low) classification of conventionality (C) using the median and novelty (N) using the tenth percentile of z-scores within an article. Thus, HCLN would denote high conventionality and low novelty with all four combinations being possible. 
\cite{uzzi_atypical_2013} observed that HCHN articles were twice as likely to be highly cited, suggesting that novel combinations of ideas flavoring a body of conventional thought were a 'nearly universal` feature for impact.

@@

A follow-up study by \citep{boyack_vs_uzzi_2014} explored the impact of discipline and journal effects on the definition of conventionality and novelty.  While their study had some methodological differences in the use of Scopus data rather than WoS data, a smaller dataset, and a $\chi^2$ calculation rather than Monte Carlo simulations to generate expected values of journal pairs, Boyack and Klavans noted strong effects from both disciplines and journals that were not reported by Uzzi et al. While they reported the trend that HCHN is more probable in highly cited papers, they observed that ``only 64.4\%  of  243  WoS  subject  categories'' in the Uzzi {\em et al.} study met the criterion of having the highest probability of hit papers in the HCHN category.  Further, they observed that journals vary widely in terms of size and influence and that 20 journals accounted for nearly 15\% of co-citations in their measurements. Lastly, they noted that three multidisciplinary journals accounted for 9.4\% of all atypical combinations.

Despite different methods used to generate expected values, both of these key preceding studies (\emph{vide supra}) measured co-citation frequencies across the scientific literature and normalized them without disciplinary constraints before subsequently analyzing disciplinary subsets. We hypothesized that normalization using expected values drawn from datasets focused in disciplinary areas, rather than all of WoS, would reduce model mis-specification.  Consequently, we used keyword searches of the scientific literature to construct citation networks themed around academic disciplines: \emph{applied physics, immunology, and metabolism}. Within these disciplinary frameworks, we calculated observed and expected co-citation frequencies using a refined random graph model and an efficient Monte Carlo simulation algorithm.

@@

\section{Materials and Methods}
\emph{Bibliographic data} We have previously developed ERNIE, an open source knowledge platform into which we parse the Web of Science (WoS) Core Collection~\citep{Keserci371955}. WoS data stored in ERNIE spans the period 1900-2019 and consists of over 72 million publications. For this study, we generated an analytical dataset from years 1985 to 2005 from ERNIE. The total number of publications in this dataset was just over 25 million publications (25,134,073), which were then stratified by year of publication. For each of these years, we further restricted analysis to publications of type Article. Since WoS data also contains incomplete references or references that point at other indexes, we also considered only those references for which there were complete records~(Table \ref{tab:summary_data}). For example, WoS data for year 2005 contained 1,753,174 publications, which after restricting to type Article and considering only those references described above resulted in 916,573 publications, 6,095,594 unique references (set of references), and 17,167,347 total references (multiset of references). Given consistent trends in the data, Table \ref{tab:summary_data}, we analyzed the two boundary years (1985 and 2005) and the mid-point (1995). We also used the number of times each of these articles was cited in the first 8 years since publication as a measure of its impact.

@@

\subsection{Model Misspecification and the Attributes of Disciplinary Context} We invoke model mis-specification arising from irrelevant references as a concern when disciplinary subsets are analyzed using a random graph model that generates expected co-citation frequencies from the full body of scientific literature. A symptom of model misspecification is that random sampling of citations associated with an article do not match with the distribution of citations across the disciplines observed in articles from that discipline. \cite{wallace_lariviere_gingras_2012} found that, often, a majority of an article's citations are from the specialty of the article, although that percentage varied among disciplines in the eight specialties they investigated: from approximately 39\% to 89\% for 2006. To address this issue, we first assembled three exemplar datasets using keyword searches for the terms `applied physics', `immunology', and `metabolism' (Materials and Methods). Using the WoS extended classificatation of 153 subject areas, we then analyzed the disciplinary composition of the references in these three datasets into the five research areas in the Web of Science (life sciences and biomedicine, physical sciences, social sciences, technology, and arts and humanities) for the years 1985, 1995, and 2005. We observed that 84.2\% of the references in the applied physics dataset were classified as belonging to the WoS physical sciences research area. Similarly, 93.7\% and 92.1\% of the references in the immunology and metabolism datasets belonged to the life sciences and biomedicine research area, indicating strongly that references are more likely belong to the same research area as the publications that cite them.

 We also analyzed the consequences of citation shuffling within a disciplinary set or all of the Web of Science.  References in publications belonging to these three datasets were classified using the WoS Extended Subjects classification of 153 subject areas and summarized as a frequency distribution. A single shuffle of the references in these datasets or all the references in the corresponding WoS year slice was performed, using either the rpcs or umsj algorithms, after which subject frequencies were computed again. The fold change in subject frequencies of references before and after shuffling was calculated for these groups using all 153 Subject categories and are summarized as boxplots in Fig \ref{fig:be}. For example, the applied physics dataset contained 1 reference labeled Genetics and Heredity . This proportion was preserved in the intra-network shuffle but increased 1496 times when the WoS background was used as a source of substitution during Monte Carlo simulation with the repcs algorithm. Similarly, with the metabolism dataset.which contained one reference labeled Philosophy that was increased to 661 after a single repcs shuffle. The data show convincingly that the disciplinary composition of references in a network is preserved when citation shuffling is conducted within it but is significantly altered when the full WoS network is used as a source of substitution. A second inference is that the choice between these two algorithms is not relevant beyond runtime advantage but that the background network used for the shuffle is critical.
